Leveraging genomics to detect trait selection from within and between species variations

### 2023-09-13 manuscript/title_page.tex
M. Bastian, T. Gaboriau and N. Salamin edits

%! BibTeX Compiler = biber
%TC:ignore
\documentclass{article}
\usepackage{caption}
\usepackage{censor}
\usepackage{xcolor, colortbl}
\definecolor{BLUELINK}{HTML}{0645AD}
\definecolor{DARKBLUELINK}{HTML}{0B0080}
\definecolor{LIGHTGREY}{gray}{0.9}
\PassOptionsToPackage{hyphens}{url}
\usepackage[colorlinks=false]{hyperref}
% for linking between references, figures, TOC, etc in the pdf document
\hypersetup{colorlinks,
    linkcolor=DARKBLUELINK,
    anchorcolor=DARKBLUELINK,
    citecolor=DARKBLUELINK,
    filecolor=DARKBLUELINK,
    menucolor=DARKBLUELINK,
    urlcolor=BLUELINK
} % Color citation links in purple
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{naturalnames}{hyperref}

\usepackage{biorxiv}
\usepackage[backend=biber,eprint=false,isbn=false,url=false,giveninits=true, style=apa,date=year]{biblatex}
\DeclareDelimFormat{nameyeardelim}{\addcomma\space}
\addbibresource{references.bib}

@@

% -- Defining colors:
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}% Definig a custom style:
\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},
    basicstyle=\ttfamily\scriptsize\bfseries,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=t,
    keepspaces=true,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2
}% -- Setting up the custom style:
\lstset{style=mystyle}
\captionsetup[table]{hypcap=false}
\captionsetup[figure]{hypcap=false}

\newcommand{\defEqual}{\stackrel{\text{def}}{=}}
\newcommand{\Multiply}{\cdot}
\newcommand{\MultiplyMatrix}{\times}
\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\newcommand{\tr}{^{\intercal}}
\newcommand{\inv}{^{-1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\text{Var}}
\DeclareMathOperator{\Cov}{\text{Cov}}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\pn}{\pi_N}
\newcommand{\ps}{\pi_S}
\newcommand{\pnps}{\pn / \ps}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\Spi}{i}
\newcommand{\Spj}{j}
\newcommand{\NbrTaxa}{n}
\newcommand{\Time}{t}
\newcommand{\Trait}{P}
\newcommand{\Heritability}{h^2}
\newcommand{\HeritabilitySpi}{\Heritability_{\Spi}}
\newcommand{\MeanTrait}{\bar{\Trait_{\Time}}}
\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
\newcommand{\Root}{0}
\newcommand{\RootTrait}{\Trait_{\Root}}
\newcommand{\VarPhy}{\Var \left[\MeanTrait\right]}
\newcommand{\VecZero}{\UniDimArray{0}}
\newcommand{\VecOne}{\UniDimArray{1}}
\newcommand{\Distance}{\BiDimArray{D}}
\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
\newcommand{\MutationRate}{\mu}
\newcommand{\SubRate}{q}
\newcommand{\NbrLoci}{L}
\newcommand{\VarPhenotype}{V_{\Trait}}
\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
\newcommand{\VarGenetic}{V_{\mathrm{A}}}
\newcommand{\VarGeneticSpi}{V_{\mathrm{A}, \Spi}}
\newcommand{\VarEnv}{V_{\mathrm{E}}}
\newcommand{\VarMutation}{V_{\mathrm{M}}}
\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
\newcommand{\RateBetween}{\sigma^2_{\mathrm{B}}}
\newcommand{\RateWhithin}{\sigma^2_{\mathrm{W}}}
\newcommand{\RateWhithinSpi}{\sigma^2_{\mathrm{W}, \Spi}}
\newcommand{\VecRateWhithin}{\UniDimArray{\RateWhithin}}
\newcommand{\EstRateBetween}{\widehat{\RateBetween}}
\newcommand{\EstRateWhithin}{\widehat{\RateWhithin}}
\newcommand{\NI}{\rho}
\newcommand{\EstNI}{\widehat{\rho}}
\newcommand{\StdSelection}{\sigma}
\newcommand{\VarSelection}{\StdSelection^2}

% Tree
\newcommand{\Nbranch}{2 \NbrTaxa - 2}
\newcommand{\WishartPostDf}{2 \NbrTaxa + 1}
\newcommand{\Ntrait}{K}
\newcommand{\contrast}{\UniDimArray{C}}
\newcommand{\Covariancematrix}{\Sigma}
\newcommand{\CovarianceMatrix}{\BiDimArray{\Covariancematrix}}
\newcommand{\Precisionmatrix}{\Omega}
\newcommand{\PrecisionMatrix}{\BiDimArray{\Precisionmatrix}}
\newcommand{\Identitymatrix}{\BiDimArray{I}}
\newcommand{\brownian}{\mathcal{B}}
\newcommand{\Brownian}{\UniDimArray{\brownian}}
\newcommand{\Scattermatrix}{\BiDimArray{A}}
\newcommand{\Multivariate}{\UniDimArray{Z}}

@@

% Including genomics, Inclusion of genomics, Linking genomics,
\title{Leveraging genomics to detect trait selection from within and between-species variations}
\rhead{\scshape Trait selection from within and between-species variations}

\author{
\large
\textbf{T. {Latrille}$^{1}$\orcidlink{0000-0002-9643-4668}, M. {Bastian}$^{2}$, T. {Gaboriau}$^{1}$\orcidlink{0000-0001-7530-2204}, N. {Salamin}$^{1}$\orcidlink{0000-0002-3963-4954}}\\
\normalsize
$^{1}$Department of Computational Biology, Université de Lausanne, Lausanne, Switzerland\\
$^{2}$Laboratoire de Biométrie et Biologie Evolutive, UMR5558, Université Lyon 1, Villeurbanne, France \\
\texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
}
\author{~}
\begin{document}

@@

% Abstract (≤ 250 words)
%TC:endignore
\begin{abstract}
    To determine whether a trait is neutrally evolving or under selection, methods have been developed using either within or between-species variation.
    However, methods using within-species variation do not integrate the changes at the macro-evolutionary scale.
    Conversely, current methods using between-species variation usually discard within-species variation, thus not accounting for processes at the micro-evolutionary scale.
    The main goal of this study is to define a neutrality index for a quantitative trait, by combining within- and between-species variation.
    This neutrality index integrates nucleotide polymorphism and divergence for normalizing trait variation.
    As such, it does not require estimation of population size nor of time of speciation for normalization.
    Our index can be used to seek deviation from the null model of neutral evolution, and test for diversifying selection.
    Applied to brain mass and body mass at the mammalian scale, we show that brain mass is under diversifying selection.
    Finally, we show that our test is not sensitive to the assumption that population sizes, mutation rates and generation time are constant across the phylogeny, and automatically adjust for it.
\end{abstract}

\keywords{Quantitative genetics \and Trait evolution \and Selection \and Phylogenetics \and Population genetics }

% Research Article (7500 words)
\section*{Introduction}\label{sec:introduction}

Determining whether a trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
Fundamentally, distinguishing neutral evolution from selection requires determining which selective regime is supported by the observed variation of traits or sequences.
The variation of phenotypes (traits) and genotypes (sequences) can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
All these systems require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\parencite{lynch_genetics_1998, walsh_evolution_2018}.

[not sure of this paragraph...]

Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait), are also mostly pleiotropic (many traits associated with a given locus), additive and mainly under stabilizing selection~\parencite{simons_population_2018, sella_thinking_2019}.
Additionally, contrasting both trait and genetic differentiation across several populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to determine the selective regime and to quantify the strength of selection acting on a trait~\parencite{martin_multivariate_2008, leinonen_qst_2013}.
For example, studies have shown that stabilizing selection is largely dominant in the evolution of gene expression in different populations~\parencite{whitehead_neutral_2006, gilad_natural_2006}. [why only example from gene expression?]
However, dominant selective regime acting on gene expression is still controversial, and neutral evolution is still debated~\parencite{signor_evolution_2018, price_detecting_2022}.

To disentangle neutral evolution and selection, trait evolution can also be observed at a larger time scale.
For example, change in mean trait value accumulates linearly with time of divergence from a sister species, and also proportionally to the trait variance~\parencite{lande_genetic_1980, turelli_heritable_1984}.
Gene expression exhibits a similar accumulation as divergence in expression accumulates faster for gene with large within-species variation~\parencite{khaitovich_neutral_2004}.
Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\parencite{walsh_evolution_2018}.

% Phylogenetics (mean trait evolution)
Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\parencite{felsenstein_phylogenies_1985}.
Under neutral evolution, the change in main trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\parencite{hansen_translating_1996}.
As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
Some deviations from the Brownian process, typically Ornstein-Uhlenbeck processes, are thus often interpreted as a signature of stabilizing selection~\parencite{catalan_drift_2019}.
Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction) is interpreted as a signature of directional selection at the phylogenetic scale~\parencite{silvestro_early_2019}.
However, studies have shown that such comparative approaches are subject to different biases~\parencite{harmon_phylogenetic_2018}.
First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\parencite{hansen_translating_1996}.
In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
Secondly, and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Those limitations, altogether with the use of mean trait estimates leaving out the variance in traits between individuals, easily generate misclassification of selection from methods at the phylogenetic scale.

% At the frontier Phylogenetics/Population-genetics
At the frontier between micro and macroevolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation on top of changes in mean trait value~\parencite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
However, in the context of detecting selection acting on a trait, within-species variation is mostly used to describe technical errors.
One notable exception used the ratio of between to within variation estimated for many traits, and compared to the average over all traits to seek deviation from this average~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
The goal of this study is to determine the evolutionary regime of a quantitative trait through the articulation of the variance between and within species while setting the threshold for neutral, stabilizing, and diversifying selection.
We propose a neutrality index for a quantitative trait articulating trait and nucleotide variation within and between species.
Importantly, our neutrality index also leverages nucleotide divergence and polymorphism in order to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between-species).
For population geneticists, our study can be seen as the macro-evolutionary generalization of Q$_\text{ST}$--F$_\text{ST}$ methods to account for phylogenetic relationships between species.
For phylogeneticists, our study can be seen as an alternative to the EVE model~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}, where we set a clear threshold for neutral evolution by reviewing theoretical literature and leveraging species' nucleotide polymorphism and divergence.
% Basically, we are testing whether the variance of means is greater than the mean of variances

\section*{Materials and Methods}\label{sec:materials-and-methods}
\subsection*{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}

Prior to developing our neutrality index, we review theoretical expectations for variations of quantitative traits and genomic sequences under neutral evolution for both within- and between-species variation.

\subsubsection*{Within-species trait variations}

For a given trait, the genetic architecture is mainly defined by the number of loci encoding the trait ($\NbrLoci$) and the random additive effect of a mutation on the trait ($a$).
New mutations are generating trait variance and the average effect of a mutation on the trait is $\RateMut = \NbrLoci \Multiply  \E [a^2]$.
At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRate$) and $\RateMut$:
\begin{gather}
    \VarMutation = 2 \MutationRate \Multiply \RateMut \label{eq:var-mutation}.
\end{gather}

While in an infinitesimal model mutations supply new genetic variants, random genetic drift depletes standing variation~\parencite{turelli_commentary_2017, barton_infinitesimal_2017, sella_thinking_2019}.
For a neutral trait at equilibrium between mutation and drift~\parencite{lynch_mutation_1998}, the additive genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$):
\begin{align}
    \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
    & = 4 \Ne \Multiply \MutationRate \Multiply \RateMut \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
\end{align}

For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is measured as the number of mutations segregating in the population divided by the length of the region.
Any segregating mutations will eventually reach fixation or extinction due to random genetic drift and $\pi$ is also at a balance between mutations and drift.
As shown in \textcite{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$):
\begin{gather}
    \pi = 4 \Ne \Multiply \MutationRate \label{eq:genetic-diversity}.
\end{gather}

We define $\RateWhithin$ as the ratio of additive genetic variance of the trait ($\VarGenetic$) over $\pi$ of any neutral genomic region of interest.
This ratio allows to remove the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RateWhithin$ as a proxy of $\RateMut$:
\begin{align}
    \RateWhithin & \defEqual \frac{\VarGenetic }{\pi}, \\
    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \RateMut}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
    & = \RateMut. \label{eq:rate-pop}
\end{align}

The additive genetic variance is also equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by narrow-sense heritability ($\Heritability$; \parencite{hill_data_2008}), which leads to $\RateWhithin$ beinf a function of $\VarPhenotype$ and $\Heritability$:
\begin{align}
    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
\end{align}

\subsubsection*{Between-species trait variations}

For a given species, we denote by $\MeanTrait$ the mean value of the trait across the individuals in the species at generation $\Time$.
If the trait is neutral and encoded by many loci as assumed by the infinitesimal model, $\MeanTrait$ evolves as a Brownian process~\parencite{felsenstein_phylogenies_1985, hansen_translating_1996}.
The variance of $\MeanTrait$ after $\Time$ generations, $\VarPhy$ is given by \parencite{hansen_translating_1996}:
\begin{align}
    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
    & = 4 \Time \Multiply \MutationRate \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
\end{align}

Moreover, for any genomic region under neutral evolution, some mutations will eventually reach fixation due to random genetic drift, resulting in a substitution of a nucleotide at the species level.
The probability of fixation ($\pfix$) of a neutral mutation is $1/2\Ne$ \parencite{kimura_probability_1962}.
We can derive the substitution rate per generation $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\parencite{mccandlish_modeling_2014}, giving:
\begin{align}
    \SubRate & = 2 \Ne \Multiply \MutationRate \Multiply \pfix, \\
    & = 2 \Ne  \Multiply \MutationRate  \Multiply \dfrac{1}{2\Ne}, \\
    & = \MutationRate. \label{eq:substitution-rate}
\end{align}
That is, if mutations are neutral, the rate of substitution within a genomic region equals the rate with which new mutations arise per generation for the same genomic region~\parencite{kimura_evolutionary_1968}.

After $\Time$ generations and assuming that no multiple substitutions occurred at the same site, the nucleotide divergence $d$, which is the fraction of the genomic region that generated a substitution, will be $\Time$ multiplied by the nucleotide substitution rate per generation ($\SubRate$):
\begin{align}
    d & = \Time \Multiply \SubRate \\
    & = \Time \Multiply \MutationRate \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
\end{align}
We define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by the nucleotide divergence of any neutral genomic region ($d$).
This ratio allows to remove the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trade, giving $\RateBetween$ as another proxy of $\RateMut$:
\begin{align}
    \RateBetween & \defEqual \frac{\VarPhy}{4 d},\\
    & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
    & = \RateMut \label{eq:rate-phy}.
\end{align}

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/fig-input-output}
    \caption{
        Between species, the mean phenotypic trait value changes along the phylogeny, allowing estimating of between-species trait variation, $\EstRateBetween$, which is normalized by nucleotide divergence.
        Within-species, the genetic variance allows estimating of within-species trait variation, $\EstRateWhithin$, which is  normalized by nucleotide diversity.
        $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
        Under neutral evolution, $\EstNI$ is expected to be equal to one.
        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\EstNI$ greater than one.
        Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow for discarding the confounding effect on mutation rate diversity, population size and divergence time.
    }
    \label{fig:methods}
\end{figure*}

\subsubsection*{Neutrality index}

The variability between either individuals or species can be obtained for both quantitative traits and genomic sequences.
At the population level, the variability of the trait between individuals can be combined with the nucleotide diversity of any neutrally evolving genomic region to obtain $\RateWhithin$, which equals $\RateMut$ if the trait is neutrally evolving (see above).
At the phylogenetic level, the variability of the mean trait value between species can be combined with the nucleotide divergence of any neutrally evolving genomic region to obtain $\RateBetween$.
Similarly, $\RateBetween=\RateMut$ if the trait is neutrally evolving and the genetic architecture of the trait has not changed along the phylogeny.
We thus have, for a neutrally evolving trait:
\begin{gather}
    \RateWhithin = \RateBetween \text{ from eq.~\ref{eq:rate-pop} and~\ref{eq:rate-phy}}, \\
    \Rightarrow \NI \defEqual \frac{\RateBetween}{\RateWhithin} = 1.
\end{gather}
We define a neutrality index $\NI = \RateBetween / \RateWhithin$ that will equal $1$ for a trait evolving neutrally.
Both $\RateBetween$ and $\RateWhithin$ can be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rate ($\MutationRate$), nor the effective population size ($\Ne$) or time of divergence ($\Time$) need to be estimated.
Moreover, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.

\subsection*{Estimate}\label{subsec:estimate}

Based on the comparative framework accounting for phylogenetic inertia~\parencite{felsenstein_phylogenies_1985, omeara_testing_2006}, we provide a Maximum likelihood estimate for $\NI$ as well as a Bayesian estimate to derive posterior probabilities that the null model of neutrality (i.e.~$\NI = 1$) is rejected.

\subsubsection*{Maximum likelihood estimate}
At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$ as nucleotide divergence in units of substitution per site) and the topology of the phylogenetic tree.
The diagonal $\Distance_{\Spi,\Spi}$ represents the total distances from the root of the tree to each taxon ($\Spi$).
The off-diagonal elements ($\Distance_{\Spi,\Spj} = \Distance_{\Spj,\Spi}$) are the distances between the root and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
The state $\RootTrait$ at the root of the tree for the trait can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values $\VecTrait$ at the tips of the tree using maximum likelihood~\parencite{omeara_testing_2006}:
\begin{gather}
    \RootTrait = \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecOne \right)\inv \Multiply \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecTrait \right), \label{eq:estimated-root-trait}
\end{gather}
where $\VecOne$ is an $\NbrTaxa \times 1$ column vector of ones.

Finally, within-species variation $\EstRateBetween$ is estimated as~\parencite{omeara_testing_2006}:
\begin{gather}
    \EstRateBetween = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrTaxa - 1}. \label{eq:estimated-rate-phy}
\end{gather}

For a given species $\Spi$ with inter-individual data available, additive genetic variance of a trait ($\VarGeneticSpi$) is the product of heritability ($\Heritability_{i}$) and phenotypic variance ($V_{\Trait, i}$).
The ratio of $\VarGeneticSpi$ over nucleotide diversity of neutrally evolving sequences ($\pi_{\Spi}$) is a sample estimate of $\RateWhithin$.
Averaged across all species, we obtain the estimate $\EstRateWhithin$ as:
\begin{gather}
    \EstRateWhithin = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa}\frac{  \VarGeneticSpi}{ \pi_{i}} = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa} \frac{  V_{\Trait, i} \Multiply \Heritability_{i}}{ \pi_{i}}. \label{eq:estimated-rate-pop}
\end{gather}

As depicted in fig.~\ref{fig:methods}, the neutrality index is estimated as:
\begin{gather}
    \EstNI = \frac{\EstRateBetween}{\EstRateWhithin}. \label{eq:estimated-NI}
\end{gather}

\subsubsection*{Bayesian estimate}
[NS: the start of this paragraph is not so clear -> to discuss]
Even though the neutrality index can be applied to any trait, we assume here that the variation between several traits is independent.
Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework~\parencite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
Here we generalize to $\Ntrait$ non-independent traits evolving along the phylogeny using the \textit{BayesCode} software~\parencite{latrille_inferring_2021}.
Their variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology (see section~S\ref{subsec:multivariate-brownian-process}).
The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait, thus corresponding to $\NI$.
With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
$\Brownian$ is sampled using a Metropolis-Hastings algorithm (acceptance-rejection), and the posterior distribution of $\CovarianceMatrix$ is Gibbs sampled.
For each trait and each species, prior on heritability ($\Heritability$) for each species is set as a uniform distribution with user-defined boundaries.
Heritability and phenotypic variance for each trait is combined with nucleotide diversity to compute $\EstRateWhithin$ for each species, and is averaged out across species (as in eq.~\ref{eq:estimated-rate-pop}).
From $\EstRateWhithin$ and $\CovarianceMatrix$, the posterior distribution of $\EstNI$ (as in eq.~\ref{eq:estimated-NI}) is obtained for each trait.
The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality (see fig.~\ref{fig:methods}), for example by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.

\subsection*{Simulation}\label{subsec:simulations}

% vG: 13.8, vP: 69.0, vE: 55.2, nbr_loci: 5000, a: 1.0, mut_rate: 1.38e-05, pop_size: 50, h2: 0.2
% pS = 0.0027577478571428568
% Tree length (dS) = 2.340579
% Tree length (My) = 1259.4490200000002
% Root age (My) = 98.9489413888889
% Root age (dS) = 0.18388820079995308
% nbr_sites_var = 27.577478571428568
% u = 1.3788739285714283e-05
% nbr_generations = 13336.114128321291
We tested the xxx [NS: explain a bit what we test here -> NI]
We simulated phenotypic values as the sum of genotypic value and an environmental effect.
The environmental effect was normally distributed with variance $\VarEnv$.
We assumed that the genotypic value was encoded by $\NbrLoci=5,000$ loci, with each locus contributing an additive effect that was normally distributed with standard deviation $a=1$ (Fig.~\ref{fig:simulator}A and section~S\ref{subsec:genotype-phenotype-map} for theoretical formulation).
We assumed a trait with a narrow-sense heritability of $\Heritability=0.2$ and computed the theoretical $\VarEnv$ accordingly (see section~S\ref{subsec:genotype-phenotype-map}).
We followed a Wright-Fisher model with mutation, selection and drift to formalize the evolutionary dynamics of the system (Fig.~\ref{fig:simulator}A\&B).
Assuming a diploid panmictic population of size $\Ne=50$ at the root of the tree, and with non-overlapping generations, we simulated explicitly each generation along an ultrametric phylogenetic tree.
For each offspring, the number of mutations was drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $, with mutation rate per generation $\MutationRate$.
From the empirical mammalian dataset (see next section), we computed an average nucleotide divergence from root to leaves of $0.18$ and average genetic diversity of $0.00276$.
We scaled parameters to fit plausible values for mammals.
We used a mutation rate of $\MutationRate=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.

The changes in log-$\MutationRate$ and log-$\Ne$ along the lineages were both modeled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRate}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, meaning standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
Additionally, the instant value of log-$\Ne$ was modeled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
The geometric Brownian motion accounted for long-term fluctuations, while the Ornstein-Uhlenbeck introduced short-term fluctuations. [NS: not clear how you combined both here]
The simulation started from an initial sequence at equilibrium at the root of the tree and, at each node, the process was split until it finally reached the leaves of the tree.
From a speciation process perspective, this is equivalent to an allopatric speciation over one generation.

Random genetic drift was introduced by re-sampling individuals at each generation, with each parent having a probability of being sampled that was proportional to its fitness ($W$).
Selection was modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype~\parencite{tenaillon_utility_2014,blanquart_epistasis_2016}.
More specifically, the fitness of an individual was given by $W = \e^{(\Trait - \lambda)^2/ \alpha}$, where $\Trait$ was the trait value of the individual, $\lambda=0.0$ was the optimal trait value, and $\alpha=0.02$ was the strength of selection.
Mutations were considered as a displacement of the phenotype in the multidimensional space.
Beneficial mutations moved the phenotype closer to the optimum, while deleterious mutations moved it further away.
Stabilizing selection was implemented by fixing the optimum phenotype to a single value ($\lambda=0.0$).
Diversifying selection was implemented by allowing the optimum phenotype to move along the phylogenetic tree as a geometric Brownian process~\parencite{hansen_stabilizing_1997} ($\lambda \sim \brownian \left(0, \sigma_{\lambda}=1.0\right)$).
Neutral evolution was implemented by fixing the fitness landscape ($W=1$), which meant that each individual had the same probability of being sampled at each generation.

>>>>

%! BibTeX Compiler = biber
%TC:ignore
\documentclass{article}
\usepackage{caption}
\usepackage{xcolor, colortbl}
\definecolor{BLUELINK}{HTML}{0645AD}
\definecolor{DARKBLUELINK}{HTML}{0B0080}
\definecolor{LIGHTGREY}{gray}{0.9}
\PassOptionsToPackage{hyphens}{url}
\usepackage[colorlinks=false]{hyperref}
% for linking between references, figures, TOC, etc in the pdf document
\hypersetup{colorlinks,
    linkcolor=DARKBLUELINK,
    anchorcolor=DARKBLUELINK,
    citecolor=DARKBLUELINK,
    filecolor=DARKBLUELINK,
    menucolor=DARKBLUELINK,
    urlcolor=BLUELINK
} % Color citation links in purple
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{naturalnames}{hyperref}

\usepackage{biorxiv}
\usepackage[backend=biber,eprint=false,isbn=false,url=false,intitle=true,style=nature,date=year]{biblatex}
\addbibresource{references.bib}

@@

% -- Defining colors:
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}% Definig a custom style:
\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},
    basicstyle=\ttfamily\scriptsize\bfseries,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=t,
    keepspaces=true,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2
}% -- Setting up the custom style:
\lstset{style=mystyle}

@@

% Including genomics, Inclusion of genomics, Linking genomics, 
\title{Leveraging genomics to detect trait selection from within and between species variations}
\rhead{\scshape Trait selection from within and between species variations}

\author{
\large
\textbf{T. {Latrille}$^{1}$\orcidlink{0000-0002-9643-4668}, M. {Bastian}$^{2}$, T. {Gaboriau}$^{1}$\orcidlink{0000-0001-7530-2204}, N. {Salamin}$^{1}$\orcidlink{0000-0002-3963-4954}}\\
\normalsize
$^{1}$Department of Computational Biology, Université de Lausanne, Lausanne, Switzerland\\
$^{2}$Laboratoire de Biométrie et Biologie Evolutive, UMR5558, Université Lyon 1, Villeurbanne, France \\
\texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
}
\begin{document}

@@

\subsection*{Acknowledgements}
 We gratefully acknowledge the help of Nicolas Lartillot, Philippe Veber, Isabela Jeronimo do Ó, Anna Marcionetti, Julien Clavel, and Daniele Silvestro for their insightful discussions and Julien Joseph for his advice and reviews concerning this manuscript.

\subsection*{Competing interests:}
The authors declare no conflicts of interest.
\subsection*{Data and materials availability:}
The data that support the findings of this study are openly available in GitHub at \href{https://github.com/ThibaultLatrille/MicMac}{github.com/ThibaultLatrille/MicMac}. Snakemake pipeline, analysis scripts and documentation are available in the repository to replicate the study.

\end{document}
